---
title: "Single-cell RNA seq  trajctories and pseudotime plots"
author: "Stefan Filges"
date: '`r format(Sys.Date())`'
output: 
  html_document:
    theme: sandstone
    highlight: tango
    code_folding: hide
    toc: true
    toc_depth: 3
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r load_libraries}
suppressMessages(library(Seurat))
suppressMessages(library(monocle))
suppressMessages(library(viridis))
suppressMessages(library(tidyverse))
```

# What does this document do?

This document uses pre-generated single cell data objects containing pseudotime/trajctory information and generates plots for data exploration and analysis. The data sets were generated using a different script, which can take a significant amount of time and is a long document containing code on many other issues. This document allows for faster and easier-to-use analysis.

# Monocle 2

## Load data

First we load a CellDataSet called cds generated with Monocle v2. This contains the expression data, cell metdata (phenotypes) as well as pseudotime values.

```{r import_data}
load("cds_monocle2.Rdata")

clustering_DEG_genes = read.csv2("monocle2_DEG_genes.csv")
```

## Including Plots

You can also embed plots, for example:

```{r monocle_v2_pseudotime, echo=FALSE}
plot1 <- monocle::plot_cell_trajectory(
  cds = cds,
  color_by = "Pseudotime"
) + scale_color_viridis_c()

plot2 <- monocle::plot_cell_trajectory(
  cds = cds, 
  color_by = "stim"
) + scale_color_manual(values = c("#ff8668","gray70", "#5e8aab", "#ffce8e"))

Seurat::CombinePlots(plots = list(plot1,plot2))
```

```{r monocle2-heatmap, eval=TRUE, message=FALSE}
to_be_tested <- rownames(clustering_DEG_genes)[order(clustering_DEG_genes$qval)][1:100]

cds_subset <- cds[as.numeric(to_be_tested),]

diff_test_res <- monocle::differentialGeneTest(
  cds_subset,
  fullModelFormulaStr = "~sm.ns(Pseudotime)"
)

sig_gene_names <- row.names(subset(diff_test_res, qval < 0.01))

monocle::plot_pseudotime_heatmap(
  cds_subset[sig_gene_names,],
  num_clusters = 4,
  cores = 4,
  cluster_rows = TRUE,
  norm_method = 'vstExprs',
  show_rownames = TRUE,
  hmcols = viridis::magma(256)
) + scale_color_viridis_d()
```

```{r}
col_vec <- cds_subset$orig.ident %>% 
  dplyr::recode(xFD = "#00BFC4", FD = "#F8766D", WT = "#7CAE00", xWT = "#C77CFF")

# red: scf FD - #F8766D
# grün: scf WT - #7CAE00
# cyan: xen FD - #00BFC4
# violett: xen WT - #C77CFF

hmap_2 <- gplots::heatmap.2(
  x = as.matrix(cds_subset[sig_gene_names,]@assayData$exprs),
  labCol = FALSE, 
  dendrogram = "both",
  trace = "none",
  ColSideColors = as.character(col_vec),
  col = viridis::magma(256),
  key = TRUE
) 

```

```{r}
cds_time <- cds_subset[,order(cds_subset$Pseudotime)]

annotation = list(
  time = cds_time$Pseudotime,
  stim = cds_time$stim
)
annotation_colors = list(
  time = viridis::viridis(256),
  stim = c("#ff8668","gray70", "#5e8aab", "#ffce8e")
)

fig1 <- NMF::aheatmap(
  x = as.matrix(exprs(cds_time[sig_gene_names,])),
  Colv = NA, Rowv = NA, border_color = NA,
  annCol = annotation,
  annColors = annotation_colors, 
  scale = 'none', filename = "aheatmap.svg"
)
```

## Plot differentially expressed genes in pseudotime

```{r}
diff_test_res <- diff_test_res[diff_test_res$qval < 0.01,]
diff_test_res <- diff_test_res[order(diff_test_res$qval),]

my_genes <- row.names(subset(fData(cds), gene_short_name %in% head(diff_test_res$gene_short_name,3))) 
cds_subset <- cds[my_genes,]

print(head(diff_test_res[,c("gene_short_name", "pval", "qval")]))
```

```{r}
# Plot by pseudotime
plot1 <- monocle::plot_genes_in_pseudotime(
  cds_subset = cds_subset,
  color_by = "Pseudotime",
  cell_size = 0.5
) + scale_color_viridis_c()

# Plot by sample
plot2 <- monocle::plot_genes_in_pseudotime(
  cds_subset = cds_subset,
  color_by = "stim",
  cell_size = 0.5
) + scale_color_manual(values = c("#ff8668","gray70", "#5e8aab", "#ffce8e"))

Seurat::CombinePlots(plots = list(plot1,plot2))
```




