---
title: "(5) Trajctories and pseudotime plots"
author: "Stefan Filges"
date: '`r format(Sys.Date())`'
output: 
  html_document:
    theme: sandstone
    highlight: tango
    code_folding: show
    toc: yes
    toc_depth: 3
    toc_float: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r load_libraries}
# Packages for main analysis, clustering, data structures etc.
suppressMessages(library(Seurat, quietly = TRUE))
suppressMessages(library(mclust, quietly = TRUE))
suppressMessages(library(biomaRt, quietly = TRUE))
suppressMessages(library(msigdbr, quietly = TRUE))
suppressMessages(library(clusterProfiler, quietly = TRUE))
suppressMessages(library(destiny, quietly = TRUE))
suppressMessages(library(SummarizedExperiment, quietly = TRUE))
suppressMessages(require(clusterExperiment, quietly = TRUE))
suppressMessages(library(MAST, quietly = TRUE))
suppressMessages(library(DESeq2, quietly = TRUE))
suppressMessages(library(sctransform, quietly = TRUE))

# Packages for data handeling and plotting
suppressMessages(library(reshape, quietly = TRUE))
suppressMessages(library(cowplot, quietly = TRUE))
suppressMessages(library(tidyverse, quietly = TRUE))
suppressMessages(library(viridis, quietly = TRUE))
suppressMessages(library(RColorBrewer, quietly = TRUE))
suppressMessages(library(VennDiagram, quietly = TRUE))
suppressMessages(library(DT, quietly = TRUE))

# Packages for pseudotime ordering
suppressMessages(library(slingshot, quietly = TRUE))
suppressMessages(library(monocle, quietly = TRUE))
suppressMessages(library(SCORPIUS, quietly = TRUE))
```

# What does this document do?

This document uses pre-generated single cell data objects containing pseudotime/trajctory information and generates plots for data exploration and analysis. The data sets were generated using a different script, which can take a significant amount of time and is a long document containing code on many other issues. This document allows for faster and easier-to-use analysis.

# Monocle 2

## Load data

First we load a CellDataSet called cds generated with Monocle v2. This contains the expression data, cell metdata (phenotypes) as well as pseudotime values.

```{r import_data}
# load object
seurat_object_save <- "~/GitHub/SingleCellScripts/data/Rdata/seurat_object_final.Rdata"
load(seurat_object_save)
load(file = '~/GitHub/SingleCellScripts/data/Rdata/cds_monocle2.Rdata')
load(file = '~/GitHub/SingleCellScripts/data/Rdata/seurat_object.Rdata')

clustering_DEG_genes = read.csv2(file = '~/GitHub/SingleCellScripts/data/gene_lists/monocle2_DEG_genes.csv')
order_genes <- read.csv2(file = '~/GitHub/SingleCellScripts/data/gene_lists/ordering_genes.csv')
```

## Cell trajectories

```{r monocle_v2_pseudotime, echo=FALSE}
plot1 <- monocle::plot_cell_trajectory(
  cds = cds,
  color_by = "Pseudotime"
) + scale_color_viridis_c()

plot2 <- monocle::plot_cell_trajectory(
  cds = cds, 
  color_by = "group"
) + scale_color_manual(
  values = c('#ff8668', 'gray70', '#5e8aab', '#ffce8e')
  )

#+ scale_color_brewer(palette = "Dark2")
#+ scale_color_manual("#ff8668","gray70","#5e8aab","#ffce8e")

plot1 + plot2
```

## Expression heatmap

We want to find genes differentially expressed across pseudotime. Instead of
testing all genes, we will limit the number of genes to be tested: (i) First
we select only genes differentially expressed in cluster with an adjusted
p value $<0.01$ and (ii) select the top 500 of those to perform the test. 

After testing we select again based on adjusted p value $<0.01$ to find the 
most significantly expressed genes along the trajectory. These are then visualised
in a heatmap with 

```{r monocle2-heatmap, eval=TRUE, message=FALSE}
clustering_genes <- clustering_DEG_genes %>% 
  dplyr::filter(use_for_ordering == TRUE) %>%
  dplyr::arrange(qval)

to_be_tested <- clustering_genes$gene_short_name[1:1500]
#to_be_tested <- order_genes$gene

cds_subset <- cds[as.character(to_be_tested),]

diff_test_res <- monocle::differentialGeneTest(
  cds_subset,
  fullModelFormulaStr = "~sm.ns(Pseudotime)"
)

sig_gene_names <- row.names(subset(diff_test_res, qval < 0.01))

heatmap <- monocle::plot_pseudotime_heatmap(
  cds_subset[sig_gene_names,],
  num_clusters = 4,
  cores = 4,
  cluster_rows = TRUE,
  norm_method = 'vstExprs',
  show_rownames = FALSE,
  hmcols = viridis::magma(256),
  return_heatmap = TRUE
)

pdf(file = '~/GitHub/SingleCellScripts/figures/heatmap_4_clusters.pdf', width = 7,height = 3)
heatmap
dev.off()

# Retrieve gene names + cluster identities
clusters <- cutree(heatmap$tree_row, k = 4)
t <- tibble(
  gene = names(clusters), 
  cluster = clusters
  ) %>%
  arrange(cluster)

readr::write_csv2(t, path = '~/GitHub/SingleCellScripts/data/gene_lists/diff_genes_over_pseudotime_top_1500_q_0.01.csv')
```

```{r}
cds_time <- cds_subset[,order(cds_subset$Pseudotime)]

# red: scf FD - #F8766D
# grün: scf WT - #7CAE00
# cyan: xen FD - #00BFC4
# violett: xen WT - #C77CFF

ggplotColours <- function(n = 6, h = c(0, 360) + 15){
  if ((diff(h) %% 360) < 1) h[2] <- h[2] - 360/n
  hcl(h = (seq(h[1], h[2], length = n)), c = 100, l = 65)
}

annotation = list(
  time = cds_time$Pseudotime,
  group = cds_time$group
)
annotation_colors = list(
  time = viridis::viridis(256),
  group =  c("#ff8668", "gray70","#5e8aab","#ffce8e")
)

NMF::aheatmap(
  x = as.matrix(exprs(cds_time[1:10,])),
  Colv = NA, 
  Rowv = NA, 
  border_color = NA,
  annCol = annotation,
  annColors = annotation_colors,
  scale = 'none',
  width = 5,
  height = 5,
  filename = "~/GitHub/SingleCellScripts/figures/aheatmap_500_genes.pdf"
)
```

## Plot differentially expressed genes in pseudotime

```{r}
diff_test_res <- diff_test_res %>% 
  dplyr::filter(qval < 0.01) %>%
  dplyr::arrange(qval)

my_genes <- c('CDKN2C', 'LAMA4', 'GAPDH', 'APOE', 'ALDOA')
cds_subset <- cds[my_genes,]

#markers_to_plot <- t %>% dplyr::group_by(cluster) %>% summarise(first = first(gene)) 
#cds_subset <- cds[markers_to_plot$first,]

# Plot by pseudotime
plot1 <- monocle::plot_genes_in_pseudotime(
  cds_subset = cds_subset,
  color_by = "Pseudotime",
  relative_expr = TRUE,
  cell_size = 0.5
) + scale_color_viridis_c() +
  theme(legend.position = 'bottom')

# Plot by sample
plot2 <- monocle::plot_genes_in_pseudotime(
  cds_subset = cds_subset,
  relative_expr = TRUE,
  color_by = "group",
  cell_size = 0.5
  ) + scale_color_manual(values = c("#ff8668", "gray70","#5e8aab","#ffce8e"))+
  theme(legend.position = 'bottom')

#+ scale_color_brewer(palette = "Dark2")
#+ scale_color_manual(values = c("#ff8668", "gray70","#5e8aab","#ffce8e"))

plot1 + plot2
```


```{r}
custom_dotplot <- function(
  egmt,
  showCategory = 20,
  x.axis = 'GeneRatio',
  font.size = 7
  ){
  data <- egmt@result %>%
    dplyr::arrange(qvalue) %>%
    dplyr::select(-geneID) %>%
    tidyr::separate(GeneRatio, c("top", "bottom"), sep = "/") %>%
    dplyr::mutate(GeneRatio = as.numeric(top)/as.numeric(bottom))
  data <- head(data, showCategory)
  data$category <- factor(data$ID, levels = rev(data$ID))
  if(x.axis == 'GeneRatio') {
    dp <- ggplot(
      data = data,
      mapping = aes(x = GeneRatio, y = category, color = -log10(qvalue))
    ) +
      geom_point(aes(size = Count)) +
      scale_color_continuous(
        low="red", high="blue",
        name = '-log10(qvalue)',
        guide=guide_colorbar(reverse=FALSE)
        ) +
      theme_minimal() +
      theme(
        axis.line = element_line(color="black", size = 0.2),
        axis.title.y = element_blank(),
        axis.text.y = element_text(size = font.size)
        ) + guides(size = guide_legend(reverse = TRUE))
  } else if (x.axis == 'qvalue'){
    dp <- ggplot(
      data = data,
      mapping = aes(x = -log10(qvalue), y = category, color = GeneRatio)
    ) +
      geom_point(aes(size = Count)) +
      scale_color_continuous(
        low="red", high="blue",
        name = 'GeneRatio',
        guide=guide_colorbar(reverse=FALSE)
        ) +
      theme_minimal() +
      theme(
        axis.line = element_line(color="black", size = 0.2),
        axis.title.y = element_blank(),
        axis.text.y = element_text(size = font.size)
        ) + guides(size = guide_legend(reverse = TRUE))
  }
  return(dp)
}

enricher_custom <- function(
  gene,
  pvalueCutoff,
  pAdjustMethod = "BH",
  universe = NULL,
  minGSSize = 10,
  maxGSSize = 500,
  qvalueCutoff = 0.2,
  USER_DATA) {
  
    gene <- as.character(unique(gene))
    qExtID2TermID <- EXTID2TERMID(gene, USER_DATA)
    qTermID <- unlist(qExtID2TermID)
    
    if (is.null(qTermID)) {
      message("--> No gene can be mapped....")
      p2e <- get("PATHID2EXTID", envir = USER_DATA)
      sg <- unlist(p2e[1:10])
      sg <- sample(sg, min(length(sg), 6))
      message("--> Expected input gene ID: ", paste0(sg, collapse = ","))
      message("--> return NULL...")
      return(NULL)
    }
    qExtID2TermID.df <- data.frame(
      extID = rep(names(qExtID2TermID),
                  times = lapply(qExtID2TermID, length)),
      termID = qTermID
    )
    
    qExtID2TermID.df <- unique(qExtID2TermID.df)
    qTermID2ExtID <- with(
      qExtID2TermID.df, split(as.character(extID),as.character(termID))
    )
    
    extID <- ALLEXTID(USER_DATA)
    
    if (missing(universe)) 
      universe <- NULL
    if (!is.null(universe)) {
      extID <- as.character(universe)
    }
    
    qTermID2ExtID <- lapply(qTermID2ExtID, intersect, extID)
    qTermID <- unique(names(qTermID2ExtID))
    termID2ExtID <- TERMID2EXTID(qTermID, USER_DATA)
    termID2ExtID <- lapply(termID2ExtID, intersect, extID)
    geneSets <- termID2ExtID
    idx <- get_geneSet_index(termID2ExtID, minGSSize, maxGSSize)
    
    if (sum(idx) == 0) {
      msg <- paste("No gene set have size >", minGSSize, "...")
      message(msg)
      message("--> return NULL...")
      return(NULL)
    }
    
    termID2ExtID <- termID2ExtID[idx]
    qTermID2ExtID <- qTermID2ExtID[idx]
    qTermID <- unique(names(qTermID2ExtID))
    
    k <- sapply(qTermID2ExtID, length)
    k <- k[qTermID]
    M <- sapply(termID2ExtID, length)
    M <- M[qTermID]
    N <- rep(length(extID), length(M))
    n <- rep(length(qExtID2TermID), length(M))
    args.df <- data.frame(
      numWdrawn = k - 1, 
      numW = M, 
      numB = N - M, 
      numDrawn = n
    )
    
    pvalues <- apply(
      args.df, 1, function(n) phyper(n[1], n[2], n[3], n[4], lower.tail = FALSE)
    )
    
    GeneRatio <- apply(
      data.frame(a = k, b = n), 1, function(x) paste(x[1], "/", x[2], sep = "", collapse = "")
    )
    
    BgRatio <- apply(
      data.frame(a = M, b = N), 1, function(x) paste(x[1], "/", x[2], sep = "", collapse = "")
    )
    
    Over <- data.frame(
      ID = as.character(qTermID), 
      GeneRatio = GeneRatio, 
      BgRatio = BgRatio, 
      pvalue = pvalues,
      stringsAsFactors = FALSE
    )
    
    p.adj <- p.adjust(Over$pvalue, method = pAdjustMethod)
    
    qobj <- tryCatch(qvalue(p = Over$pvalue, 
                            lambda = 0.05,
                            pi0.method = "bootstrap"),
                     error = function(e) NULL)
    
    if (class(qobj) == "qvalue") {
      qvalues <- qobj$qvalues
    }
    else {
      qvalues <- NA
    }
    
    geneID <- sapply(qTermID2ExtID, function(i) paste(i, collapse = "/"))
    geneID <- geneID[qTermID]
    
    Over <- data.frame(
      Over,
      p.adjust = p.adj,
      qvalue = qvalues,
      geneID = geneID, 
      Count = k, 
      stringsAsFactors = FALSE
    )
    
    Description <- TERM2NAME(qTermID, USER_DATA)
    
    if (length(qTermID) != length(Description)) {
      idx <- qTermID %in% names(Description)
      Over <- Over[idx, ]
    }
    
    Over$Description <- Description
    nc <- ncol(Over)
    Over <- Over[, c(1, nc, 2:(nc - 1))]
    Over <- Over[order(pvalues), ]
    Over$ID <- as.character(Over$ID)
    Over$Description <- as.character(Over$Description)
    row.names(Over) <- as.character(Over$ID)
    
    x <- new("enrichResult", result = Over, pvalueCutoff = pvalueCutoff, 
             pAdjustMethod = pAdjustMethod, qvalueCutoff = qvalueCutoff, 
             gene = as.character(gene), universe = extID, geneSets = geneSets, 
             organism = "UNKNOWN", keytype = "UNKNOWN", ontology = "UNKNOWN", 
             readable = FALSE)
    
    return(x)
}

enricher_edit<-function(gene, pvalueCutoff = 0.05, pAdjustMethod = "BH", universe, 
                         minGSSize = 10, maxGSSize = 500, qvalueCutoff = 0.2, TERM2GENE, 
                         TERM2NAME = NA) 
{
  USER_DATA <- build_Anno(TERM2GENE, TERM2NAME)
  enricher_custom(gene = gene, pvalueCutoff = pvalueCutoff, 
                  pAdjustMethod = pAdjustMethod, universe = universe, minGSSize = minGSSize, 
                  maxGSSize = maxGSSize, qvalueCutoff = qvalueCutoff, USER_DATA = USER_DATA)
}

body(enricher)<-body(enricher_edit)

environment(enricher_custom)<-environment(clusterProfiler:::enricher_internal)

enrichment_plot <- function(data, cluster){
  gene <- data %>% dplyr::filter(cluster == cluster)
  
  kegg <- read.gmt("~/GitHub/SingleCellScripts/data/MSigDB/c2.cp.kegg.v7.1.symbols.gmt")
  cgp <- read.gmt("~/GitHub/SingleCellScripts/data/MSigDB/c2.cgp.v7.1.symbols.gmt")
  
  hgnc_symbols <- read.csv("~/GitHub/SingleCellScripts/data/MSigDB/hgnc.txt", sep = '\t', header=TRUE)
  hgnc_symbols <- hgnc_symbols$Approved.symbol

  egmt <- enricher(
    gene$gene,
    TERM2GENE=cgp,
    pvalueCutoff = 0.1,
    pAdjustMethod = 'fdr',
    minGSSize = NA,
    maxGSSize = NA,
    qvalueCutoff = 0.05,
    universe = hgnc_symbols
  )
  cgp_dotplot <- custom_dotplot(egmt, x.axis = "qvalue")
  
  egmt <- enricher(
    gene$gene,
    TERM2GENE=kegg,
    pvalueCutoff = 0.1,
    pAdjustMethod = 'fdr',
    minGSSize = NA,
    maxGSSize = NA,
    qvalueCutoff = 0.05,
    universe = hgnc_symbols
  )
  kegg_dotplot <- custom_dotplot(egmt, x.axis = "qvalue")
  
  plot <- cgp_dotplot + kegg_dotplot
  return(plot)
}


pdf(file = paste("enrichment_cluster_1",".pdf", sep=""),width = 12,height = 6)
enrichment_plot(data = t, cluster = 1)
dev.off()
  
pdf(file = paste("enrichment_cluster_2",".pdf", sep=""),width = 12,height = 6)
enrichment_plot(2)
dev.off()
  
pdf(file = paste("enrichment_cluster_3",".pdf", sep=""),width = 12,height = 6)
enrichment_plot(3)
dev.off()
  
pdf(file = paste("enrichment_cluster_4",".pdf", sep=""),width = 12,height = 6)
enrichment_plot(4)
dev.off()
```



